Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-06-18 icfp24paper/paper.tex
icfp24 final version folder

\author{First Last}
\affiliation{%
  \institution{Institution}
  \country{Country}
  \city{City}
}

@@

We use \textbf{unstaging} to refer to the process of extracting object code from
CFTT programs, by evaluating all metaprograms in splices. This term has been
sporadically used in the literature, e.g.\ by Odersky and Rompf
\cite{DBLP:journals/cacm/RompfO12} and also by Choi et
al.\ \cite{DBLP:conf/popl/ChoiAYT11} with a somewhat different meaning in a
multi-stage context. The main precursor to this paper used ``staging'' instead
of our ``unstaging'' \cite{staged2ltt}; we use the latter because the former
conflicts with other common usages of ``staging''.

@@

Axioms clearly block computation, so they are incompatible with the mentioned
notion of soundness. But we are only interested in equations which hold
definitionally during unstaging. Hence, we can do the following: first, we erase
all identity proofs and their transports from CFTT programs, then we proceed
with unstaging as in \cite{staged2ltt}. This erasure can be formalzied as a
syntactic translation \cite{next700} from CFTT to CFTT extended with the
equality reflection principle, which says that propositional equality implies
definitional equality (see e.g.\ \cite{hofmann95conservativity}). Thus,
soundness of staging for CFTT holds up to erasure of identity proofs.

@@

A principled solution to the previous issue is to write code generators in
continuation-passing style, as first proposed by Bondorf
\cite{DBLP:conf/lfp/Bondorf92}. This can be structured as a continuation monad,
and later work used explicit monadic notation for it
\cite{DBLP:conf/emsoft/KiselyovST04,DBLP:journals/scp/CaretteK11,DBLP:conf/pepm/SwadiTKP06}. Our
definition is as follows:
\begin{alignat*}{3}
  & \newtype \Gen\,(\vA : \MTy) = \Gen\,\{\unGen : \{\vR : \Ty\} \to (\vA \to \Up \vR) \to \Up \vR\}
\end{alignat*}
This is a monad in $\MTy$ in a standard sense:
\begin{alignat*}{3}
  &\mathrlap{\instance \Monad\,\Gen\,\where}\\
  &\ind \return \va        &&= \Gen \dlr \lam \vk.\,\vk\,\va\\
  &\ind \mit{ga} \bind \vf &&= \Gen \dlr \lam \vk.\,\unGen\,\mit{ga}\,(\lam \va.\,\unGen\,(\vf\,\va)\,\vk))
\end{alignat*}
In \cite{DBLP:conf/emsoft/KiselyovST04},\cite{DBLP:journals/scp/CaretteK11} and
\cite{DBLP:conf/pepm/SwadiTKP06}, the answer type is a parameter to the
continuation monad, while we have it as polymorphic. We have found that
polymorphic answer types are more convenient, because we do not have to
anticipate the type of the code output when are defining a code generator.
Hence, we can ``run'' actions as follows:
\begin{alignat*}{3}
  &\runGen : \Gen\,(\Up \vA) \to \Up \vA\\
  &\runGen\,\vma = \unGen\,\vma\,\id
\end{alignat*}
From now on, we reuse Haskell-style type classes and $\mbf{do}$-notation in
CFTT. We will use type classes in an informal way, without precisely specifying
how they work. However, type classes are used in essentially the same way in the
Agda and Haskell implementations, with modest technical differences.  From the
$\Monad$ instance, the $\Functor$ and $\Applicative$ instances can be also
derived. We reuse $({\fmap})$ and $({\ap})$ for applicative notation as well.

@@

First, following the style of Haskell's monad transformer library \texttt{mtl}
\cite{mtl}, we define a class for monads that support code generation.
\begin{alignat*}{3}
&\tyclass\,\Monad\,\vM \RA \MonadGen\,\vM\,\where\\
&\ind \liftGen : \Gen\,\vA \to \M\,\vA
\end{alignat*}
We also redefine $\gen$ and $\genRec$ to work in any $\MonadGen$, so from now on
we have:
\begin{alignat*}{3}
 &\gen   &&: \MonadGen\,\vM \RA \Up \vA \to \vM\,(\Up \vA)\\
 &\genRec &&: \MonadGen\,\vM \RA (\Up \vA \to \Up \vA) \to \vM\,(\Up \vA)
\end{alignat*}
In the rest of this paper, we shall only mention $\ReaderT$, $\MaybeT$ and
$\StateT$ as monad transformers. For all of these, we can define the $\MonadGen$
instance simply by $\liftGen = \lift \circ \liftGen$.

@@

We demonstrate this for lists. An object-level $\mbf{case}$ on lists introduces
two points where code generation can continue. We define a metatype which gives
us a ``view'' on these points:
\[ \data \msf{SplitList}\,\vA = \Nil'\,|\,\Cons'\,(\Up \vA)\,(\Up (\List\,\vA)) \]
We can generate code for a case split, returning a view on it:
\begin{alignat*}{3}
  &\mit{split} : \Up (\List\,\vA) \to \Gen\,(\msf{SplitList}\,\vA)\\
  &\mit{split}\,\vas = \Gen \dlr \lam \vk.\,\qt{\case \spl \vas\,\of \Nil \to \spl(\vk\,\Nil'); \Cons\,\va\,\vas \to \spl(\vk\,(\Cons'\,\qt{a}\,\qt{\vas}))}
\end{alignat*}
Now, in any $\MonadGen$, assuming $\vas : \Up(\List\,\vA)$, we may write
\begin{alignat*}{3}
  &\mdo \{\mit{sp} \fro \liftGen\;(\!\mit{split}\,\vas);(\case \mit{sp}\,\of \Nil' \to ...;\Cons'\,\va\,\vas \to ...)\}
\end{alignat*}
This can be generalized to splitting on any object value. In the Agda and
Haskell implementations, we overload $\mit{split}$ with a class similar to the
following:
\begin{alignat*}{3}
  & \hspace{-2em}\mathrlap{\tyclass\,\mit{Split}\,(\vA : \VTy)\,\where}\\
  & \hspace{-2em}\ind \mit{SplitTo} &&: \MTy \\
  & \hspace{-2em}\ind \mit{split}   &&: \Lift \vA \to \Gen\,\mit{SplitTo}
\end{alignat*}
In a native implementation of CFTT it may make sense to extend do-notation, so
that we elaborate $\mbf{case}$ on object values to an application of the
appropriate splitting function. We adopt this in the rest of the paper, so
when working in a $\MonadGen$, we can write $\case \vas\,\of \Nil \to ...;
\Cons\,\va\,\vas \to ...$, binding $\va : \Up \vA$ and $\vas : \Up(\List\,\vA)$
in the $\Cons$ case.

@@

There is a deficiency in our library so far. Consider:
\begin{alignat*}{3}
  & \vf : \Bool \to \StateT\,\Nat\,(\MaybeT\,\Identity)\,()\\
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
  & \ind\modify'\,(\lambda\,x.\,\qt{\spl x + \spl x}))
\end{alignat*}
This is unstaged to the following (omitting newtype wrappers):
\begin{alignat*}{3}
  & \vf\,\vb\,\vs := \case\,\vb\,\of \\
  & \ind \True\hspace{0.2em} \to \letdef \vs := 10; \letdef \vs := \vs + \vs; \Just\,((),\vs)\\
  & \ind \False \to \letdef \vs := 20; \letdef \vs := \vs + \vs; \Just\,((),\vs)
\end{alignat*}
Notice that the final state modification gets inlined into both branches. This
duplication follows from the definition of monadic binding in $\Gen$ and the
$\mit{split}$ function in the desugaring of $\mbf{case}$. Code generation is
continued in both branches with the same action. If we have multiple Boolean
$\mbf{case}$ splits sequenced after each other, that yields exponential code
size. A possible fix is to let-bind the branching action:
\begin{alignat*}{3}
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \msf{act} \fro \gen \dlr \mdown \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20} \\
  & \ind \mup\,\msf{act} \\
  & \ind\modify\,(\lambda\,x.\,\qt{\spl x + \spl x}))
\end{alignat*}
However, this yields suboptimal code:
\begin{alignat*}{3}
  & \vf\,\vb\,\vs :=\\
  & \ind \letdef\,\msf{act}\,\vs := \case\,\vb\,\of \True\hspace{0.2em}\to \letdef \vs := 10; \Just((), \vs)\\
  & \ind\hspace{8.5em}                               \False \to \letdef \vs := 20; \Just((), \vs);\\
  & \ind \case \msf{act}\,\vs\,\of\\
  & \ind\ind \Just\,(\_,\,\vs) \to \letdef \vs := \vs + \vs; \Just\,((),\,\vs)\\
  & \ind\ind \Nothing\hspace{0.6em} \to \Nothing
\end{alignat*}
This solution is fairly good when we only have $\State$ or $\Reader$ effects, where
$\mdown$ only introduces a runtime pair constructor, and it is feasible to
compile object-level pairs as unboxed data, without overheads. However, for
$\Maybe$, $\mdown$ introduces a runtime $\Just$ or $\Nothing$, and $\mup$
introduces a runtime case split. A better solution would be to introduce two
let-bound join points before the offending $\mbf{case}$, one for returning a
$\Just$ and one for returning $\Nothing$, but fusing away the actual runtime
constructors:
\begin{alignat*}{3}
  & \vf\,\vb\,\vs :=\\
  & \ind \letdef\,\msf{joinNothing}\,\vs := \Nothing;\\
  & \ind \letdef\,\msf{joinJust}\,\hspace{1.8em}\vs := \letdef \vs := \vs + \vs; \Just\,((),\,\vs);\\
  & \ind \case \msf{b}\,\of\\
  & \ind \ind \True\hspace{0.2em}  \to \letdef \vs := 10; \msf{joinJust}\,\vs\\
  & \ind \ind \False \to \letdef \vs := 20; \msf{joinJust}\,\vs
\end{alignat*}
Here, $\msf{joinNothing}$ happens to be dead code, but it is easy to clean that up in
downstream compilation. Such ``fused'' returns are possible whenever we have a
$\Gen\,\vA$ action at the bottom of the transformer stack, such that $\vA$ is
isomorphic to a meta-level finite sum of value types. Recall that $\Gen\,\vA$ is
defined as $\{\vR : \VTy\} \to (\vA \to \Up \vR) \to \Up \vR$. Here, if $\vA$ is
a finite sum, we can rearrange $\vA \to \Up \vR$ to a finite product of
functions.

@@

How do we generate object code from streams? The state $\vS$ is given as a
finite sums of products, but the sums and the products are on the meta level, so
we cannot directly use $\vS$ in object code. Similarly as in the the treatment
of join points, we tabulate the $\vS \to \Gen\,\Step\,\vS\,\vA$ transition
function to a product of functions. However, these functions need to be
\emph{mutually recursive}, since it is possible to transition from any state to
any other state, and each such transition is represented as a function call.
This problem of generating well-typed mutual blocks was addressed by Yallop and
Kiselyov in \cite{DBLP:conf/pepm/YallopK19}. In contrast to this work, which
used control effects and mutable references in MetaOCaml, we present a solution
that does not use side effects in the metalanguage.

@@

Can we have a list-like $\Monad$ instance for streams, with singleton streams
for $\return$ and $\concatMap$ for binding? This is not possible. Aiming at
\[ \concatMap : (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
the $\vA \to \Pull\,\vB$ function can contain an infinite number of different
machine state types, which cannot be represented in a finite amount of object
code. Here by ``infinite'' we mean the notion that is internally available in
the meta type theory. For instance, we can define a $\Nat_\M \to \Pull\,\vB$
function which for each $\vn : \Nat_\M$ produces a concatenation of $\vn$
streams. Hence, we shall have the following function instead:
\[ \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB \]
The idea is the following: if $\USOP$ is closed under $\Sigma$-types, we can directly
define this $\concatMap$, by taking the appropriate dependent sum of the $\vA
\to \USOP$ family of machine states, which we extract from the $\vA \to
\Pull\,\vB$ function. Let us write $\Sigma\,\vA\,\vB : \MTy$ for dependent sums,
for $\vA : \MTy$ and $\vB : \vA \to \MTy$, and reuse $({,})$ for pairing.  We
also use the following field projection functions: $\mit{projS} : \Pull\,\vA \to
\MTy$, $\mit{projSeed} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas$, and
$\mit{projStep} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas \to
\Gen\,(\Step\,(\mit{projS}\,\vas)\,\vA)$.  For now, we assume that the following
instance exists:
\[ \instance (\IsSOP\,\vA,\,\{\va : \vA\} \to \IsSOP\,(\vB\,\va)) \RA \IsSOP\,(\Sigma\,\vA\,\vB) \]
Above, $\{\va : \vA\} \to \IsSOP\,(\vB\,\va)$ is a universally quantified
instance constraint; this can be also represented in Agda. The definition of
$\concatMap$ is as follows.
\begin{alignat*}{3}
  & \mathrlap{\concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB }\\
  & \mathrlap{\concatMap\,\{\vA\}\,\{\vB\}\,\vf\,(\Pull\,\vS\,\seed\,\step) = }\\
  & \mathrlap{\ind \Pull\,(\vS,\,\Maybe\,(\Sigma\,\vA\,(\mit{projS} \circ \vf)))\,(({,}) \fmap \seed\,\ap\,\return\,\Nothing) \dlr \lam \case }\\
  & \mathrlap{\ind \ind (\vs,\,\Nothing) \to \step\,\vs \bind \lam \case }\\
  & \ind \ind \ind \ind \Stop                 &&\to \return \Stop\\
  & \ind \ind \ind \ind \Skip\,\vs            &&\to \return \dlr \Skip\,(\vs,\, \Nothing)\\
  & \ind \ind \ind \ind \Yield\,\va\,\vs      &&\to \mdo \{\vs' \fro \mit{projSeed}\,(\vf\,\va); \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\}\\
  & \mathrlap{\ind \ind (\vs,\,\Just\,(\va,\,\vs')) \to \mit{projStep}\,(\vf\,\va)\,\vs' \bind \lam \case }\\
  & \ind \ind \ind \ind \Stop                &&\to \return \dlr \Skip\,(\vs,\,\Nothing) \\
  & \ind \ind \ind \ind \Skip\,\vs'          &&\to \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\\
  & \ind \ind \ind \ind \Yield\,\vb\,\vs'    &&\to \return \dlr \Yield\,\vb\,(\vs,\, \Just\,(\va,\,\vs'))
\end{alignat*}
Here, $\Nothing$ marks the states where we are in the ``outer'' loop, running
the $\Pull\,\vA$ stream until we get its next value. $\Just$ marks the states of
the ``inner'' loop, where we have a concrete $\va : \vA$ value and we run the
$(\vf\,\va)$ stream until it stops. In the inner loop, the machine state type
depends on the $\va : \vA$ value, hence the need for $\Sigma$. How do we get
$\IsSOP$ for $\Sigma$? The key observations are:
\begin{itemize}
 \item Metaprograms cannot inspect the structure of object terms.
 \item Object types do not depend on object terms.
\end{itemize}
Hence, we expect that during staging, every $\vf : \Lift\,\vA \to \VTy$ function
has to be constant. This is actually true in the staging semantics of
CFTT. There, all metafunctions are stable under object-level parallel
substitutions. Also, object types are untouched by substitution. Hence, a
straightforward unfolding of semantic definitions validates the constancy of
$\vf$.

Generally speaking, in the semantics, every function whose domain is a product
of object types and whose codomain is a constant presheaf, is a constant
function. We may call these constancy statements \emph{generativity axioms},
since they reflect the inability of metaprograms to inspect terms, and
``generativity'' often refers to this property in the staged compilation
literature.

Let us write $\Uprod = \List\,\VTy$ and $\Elprod : \Uprod \to \MTy$ for a
universe of finite products of value types. Concretely, in CFTT and the Agda
implementation, we assume the following:
\\\\
\noindent \textbf{Axiom (generativity).}\textit{ For every $\vf : \Elprod\,\vA \to \USOP$ and $\va,\,\va' : \Elprod\,\vA$,
we have that $\vf\,\va = \vf\,\va'$.}
\\\\
We remark that there is no risk of staging getting stuck on this axiom, because
propositional equality proofs get erased, as we described in Section
\ref{sec:semantics-of-staging}.

%% We also remark that although generativity is inconsistent with inspecting the
%% structure of object terms, it is consistent with inspecting the structure of
%% object \emph{types}.

@@

Nonetheless, it seems to be possible and useful to have a
system which simultaneously supports intensional analysis through contextual
types and the 2LTT-style implicit contexts.

@@

We believe that a programming language in the style of this paper would be
highly useful, especially in high-performance functional programming and
domain-specific programming. In particular, the pipeline for using and compiling
monads could look like the following.
\begin{itemize}
\item Users write definitions in a style similar to Haskell, without quotes,
      splices, $\mup$-s and $\mdown$-s, only marking stages in type annotations and
      in let-definitions, with $:=$ and $=$.  Storing monadic actions at runtime
      requires an explicit boxing operation that yields a closure type.
\item Type- and stage-directed elaboration adds the missing operations, and also desugars
      case splitting using $\join$ and the underlying splitting function.
\item In the downstream compiler, fast \& conservative optimization passes are
      sufficient, since abstraction overheads are already eliminated by staging. Eliminating
      dead code and unused arguments would be important, since we have not yet addressed
      this through staging.
\item A systematic way to de-duplicate code would be also needed, probably already
      during staging.
\end{itemize}

@@

\subsection{Generativity}

>>>>

\author{András Kovács}
\orcid{0000-0002-6375-9781}
\email{andrask@chalmers.se}
\affiliation{%
  \institution{University of Gothenburg}
  \country{Sweden}
  \city{Gothenburg}
}

@@

We use \textbf{unstaging} to refer to the process of extracting object code from
CFTT programs, by evaluating all metaprograms in splices. This term has been
sporadically used in the literature, see
e.g.\ \cite{DBLP:journals/cacm/RompfO12}, or \cite{DBLP:conf/popl/ChoiAYT11},
with a somewhat different meaning in a multi-stage context. The main precursor
to this paper used ``staging'' instead of our ``unstaging'' \cite{staged2ltt};
we use the latter because the former conflicts with other common usages of
``staging''.

@@

Axioms clearly block computation, so they are incompatible with the mentioned
notion of soundness. But we are only interested in equations which hold
definitionally during unstaging. Hence, we can do the following: first, we erase
all identity proofs and their transports from CFTT programs, then we proceed
with unstaging as in \cite{staged2ltt}. This erasure can be formalized as a
syntactic translation \cite{next700} from CFTT to CFTT extended with the
equality reflection principle, which says that propositional equality implies
definitional equality (see e.g.\ \cite{hofmann95conservativity}). Thus,
soundness of staging for CFTT holds up to erasure of identity proofs.

@@

A principled solution to the previous issue is to write code generators in
continuation-passing style, as first proposed by Bondorf
\cite{DBLP:conf/lfp/Bondorf92}. This can be structured as a continuation monad,
and later works used explicit monadic notation for it
\cite{DBLP:conf/emsoft/KiselyovST04,DBLP:journals/scp/CaretteK11,DBLP:conf/pepm/SwadiTKP06}. Our
definition is as follows:
\begin{alignat*}{3}
  & \newtype \Gen\,(\vA : \MTy) = \Gen\,\{\unGen : \{\vR : \Ty\} \to (\vA \to \Up \vR) \to \Up \vR\}
\end{alignat*}
This is a monad in $\MTy$ in a standard sense:
\begin{alignat*}{3}
  &\mathrlap{\instance \Monad\,\Gen\,\where}\\
  &\ind \return \va        &&= \Gen \dlr \lam \vk.\,\vk\,\va\\
  &\ind \mit{ga} \bind \vf &&= \Gen \dlr \lam \vk.\,\unGen\,\mit{ga}\,(\lam \va.\,\unGen\,(\vf\,\va)\,\vk))
\end{alignat*}
In \cite{DBLP:conf/emsoft/KiselyovST04},\cite{DBLP:journals/scp/CaretteK11} and
\cite{DBLP:conf/pepm/SwadiTKP06}, the answer type is a parameter to the
continuation monad, while we have it as polymorphic. We have found that
polymorphic answer types are more convenient, because we do not have to
anticipate the type of the code output when are defining a code generator. We
can ``run'' actions as follows:
\begin{alignat*}{3}
  &\runGen : \Gen\,(\Up \vA) \to \Up \vA\\
  &\runGen\,\vma = \unGen\,\vma\,\id
\end{alignat*}
From now on, we reuse Haskell-style type classes and $\mbf{do}$-notation in
CFTT. We will use type classes in an informal way, without precisely specifying
how they work. However, type classes are used in essentially the same way in the
Agda and Haskell implementations, with modest technical differences.  From the
$\Monad$ instance, the $\Functor$ and $\Applicative$ instances can be also
derived. We reuse $({\fmap})$ and $({\ap})$ for applicative notation as well.

@@

First, following the style of Haskell's monad transformer library \texttt{mtl}
\cite{mtl}, we define a class for monads that support code generation.
\begin{alignat*}{3}
&\tyclass\,\Monad\,\vM \RA \MonadGen\,\vM\,\where\\
&\ind \liftGen : \Gen\,\vA \to \M\,\vA
\end{alignat*}
We also redefine $\gen$ and $\genRec$ to work in any $\MonadGen$, so from now on
we have:
\begin{alignat*}{3}
 &\gen   &&: \MonadGen\,\vM \RA \Up \vA \to \vM\,(\Up \vA)\\
 &\genRec &&: \MonadGen\,\vM \RA (\Up \vA \to \Up \vA) \to \vM\,(\Up \vA)
\end{alignat*}
In the rest of this paper, we only present $\ReaderT$, $\MaybeT$ and
$\StateT$ as monad transformers. For all of these, we can define the $\MonadGen$
instance simply by $\liftGen = \lift \circ \liftGen$.

@@

We demonstrate this for lists. An object-level $\mbf{case}$ on lists introduces
two points where code generation can continue. We define a metatype which gives
us a ``view'' on these points:
\[ \data \msf{SplitList}\,\vA = \Nil'\,|\,\Cons'\,(\Up \vA)\,(\Up (\List\,\vA)) \]
We can generate code for a case split, returning a view on it:
\begin{alignat*}{3}
  &\mit{split} : \Up (\List\,\vA) \to \Gen\,(\msf{SplitList}\,\vA)\\
  &\mit{split}\,\vas = \Gen \dlr \lam \vk.\,\qt{\case \spl \vas\,\of \Nil \to \spl(\vk\,\Nil'); \Cons\,\va\,\vas \to \spl(\vk\,(\Cons'\,\qt{\va}\,\qt{\vas}))}
\end{alignat*}
Now, in any $\MonadGen$, assuming $\vas : \Up(\List\,\vA)$, we may write
\begin{alignat*}{3}
  &\mdo \{\mit{sp} \fro \liftGen\;(\!\mit{split}\,\vas);(\case \mit{sp}\,\of \Nil' \to ...;\Cons'\,\va\,\vas \to ...)\}
\end{alignat*}
This can be generalized to splitting on any object value. In the Agda and
Haskell implementations, we overload it with a class similar to the following:
\begin{alignat*}{3}
  & \hspace{-2em}\mathrlap{\tyclass\,\mit{Split}\,(\vA : \VTy)\,\where}\\
  & \hspace{-2em}\ind \mit{SplitTo} &&: \MTy \\
  & \hspace{-2em}\ind \mit{split}   &&: \Lift \vA \to \Gen\,\mit{SplitTo}
\end{alignat*}
In a native implementation of CFTT it may make sense to extend do-notation, so
that we elaborate $\mbf{case}$ on object values to an application of the
appropriate splitting function. We adopt this in the rest of the paper, so
when working in a $\MonadGen$, we can write $\case \vas\,\of \Nil \to ...;
\Cons\,\va\,\vas \to ...$, binding $\va : \Up \vA$ and $\vas : \Up(\List\,\vA)$
in the $\Cons$ case.

@@

There is a deficiency in our library so far. Consider:
\begin{alignat*}{3}
  & \vf : \Bool \to \StateT\,\Nat\,(\MaybeT\,\Identity)\,()\\
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
  & \ind\modify'\,(\lambda\,x.\,\qt{\spl x + \spl x}))
\end{alignat*}
This is unstaged to the following (omitting newtype wrappers):
\begin{alignat*}{3}
  & \vf\,\vb\,\vs := \case\,\vb\,\of \\
  & \ind \True\hspace{0.2em} \to \letdef \vs := 10; \letdef \vs := \vs + \vs; \Just\,((),\vs)\\
  & \ind \False \to \letdef \vs := 20; \letdef \vs := \vs + \vs; \Just\,((),\vs)
\end{alignat*}
Notice that the final state modification gets inlined into both branches. This
duplication follows from the definition of monadic binding in $\Gen$ and the
$\mit{split}$ function in the desugaring of $\mbf{case}$. Code generation is
continued in both branches with the same action. If we have multiple Boolean
$\mbf{case}$ splits sequenced after each other, that yields exponential code
size. A possible fix is to let-bind the branching action:
\begin{alignat*}{3}
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \msf{act} \fro \gen \dlr \mdown \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20} \\
  & \ind \mup\,\msf{act} \\
  & \ind\modify\,(\lambda\,x.\,\qt{\spl x + \spl x}))
\end{alignat*}
However, this yields suboptimal code:
\begin{alignat*}{3}
  & \vf\,\vb\,\vs :=\\
  & \ind \letdef\,\msf{act}\,\vs := \case\,\vb\,\of \True\hspace{0.2em}\to \letdef \vs := 10; \Just\,((), \vs)\\
  & \ind\hspace{8.5em}                               \False \to \letdef \vs := 20; \Just\,((), \vs);\\
  & \ind \case \msf{act}\,\vs\,\of\\
  & \ind\ind \Just\,(\_,\,\vs) \to \letdef \vs := \vs + \vs; \Just\,((),\,\vs)\\
  & \ind\ind \Nothing\hspace{0.6em} \to \Nothing
\end{alignat*}
This solution is fairly good when we only have $\State$ or $\Reader$ effects, where
$\mdown$ only introduces a runtime pair constructor, and it is feasible to
compile object-level pairs as unboxed data, without overheads. However, for
$\Maybe$, $\mdown$ introduces a runtime $\Just$ or $\Nothing$, and $\mup$
introduces a runtime case split. A better solution would be to introduce two
let-bound join points before the offending $\mbf{case}$, one for returning a
$\Just$ and one for returning $\Nothing$, but fusing away the actual runtime
constructors:
\begin{alignat*}{3}
  & \vf\,\vb\,\vs :=\\
  & \ind \letdef\,\msf{joinNothing}\,\vs := \Nothing;\\
  & \ind \letdef\,\msf{joinJust}\,\hspace{1.8em}\vs := \letdef \vs := \vs + \vs; \Just\,((),\,\vs);\\
  & \ind \case \msf{b}\,\of\\
  & \ind \ind \True\hspace{0.2em}  \to \letdef \vs := 10; \msf{joinJust}\,\vs\\
  & \ind \ind \False \to \letdef \vs := 20; \msf{joinJust}\,\vs
\end{alignat*}
Here, $\msf{joinNothing}$ happens to be dead code, but it is easy to clean up in
downstream compilation. Such ``fused'' returns are possible whenever we have a
$\Gen\,\vA$ action at the bottom of the transformer stack, such that $\vA$ is
isomorphic to a meta-level finite sum of value types. Recall that $\Gen\,\vA$ is
defined as $\{\vR : \VTy\} \to (\vA \to \Up \vR) \to \Up \vR$. Here, if $\vA$ is
a finite sum, we can rearrange $\vA \to \Up \vR$ to a finite product of
functions.

@@

How do we generate object code from streams? The state $\vS$ is given as a
finite sums of products, but the sums and the products are on the meta level, so
we cannot directly use $\vS$ in object code. Similarly as in the the treatment
of join points, we tabulate the $\vS \to \Gen\,\Step\,\vS\,\vA$ transition
function to a product of functions. However, these functions need to be
\emph{mutually recursive}, since it is possible to transition from any state to
any other state, and each such transition is represented as a function call.
This problem of generating well-typed mutual blocks was addressed in
\cite{DBLP:conf/pepm/YallopK19}. In contrast to this work, which used control
effects and mutable references in MetaOCaml, we present a solution that does not
use side effects in the metalanguage.

@@

Can we have a list-like $\Monad$ instance for streams, with singleton streams
for $\return$ and $\concatMap$ for binding? This is not possible. Aiming at
\[ \concatMap : (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
the $\vA \to \Pull\,\vB$ function can contain an infinite number of different
machine state types, which cannot be represented in a finite amount of object
code. Here by ``infinite'' we mean the notion that is internally available in
the meta type theory. For instance, we can define a $\Nat_\M \to \Pull\,\vB$
function which for each $\vn : \Nat_\M$ produces a concatenation of $\vn$
streams. Hence, we shall have the following function instead:
\[ \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB \]
The idea is the following: if $\USOP$ is closed under $\Sigma$-types, we can directly
define this $\concatMap$, by taking the appropriate dependent sum of the $\vA
\to \USOP$ family of machine states, which we extract from the $\vA \to
\Pull\,\vB$ function. Let us write $\Sigma\,\vA\,\vB : \MTy$ for dependent sums,
for $\vA : \MTy$ and $\vB : \vA \to \MTy$, and reuse $({,})$ for pairing.  We
also use the following field projection functions: $\mit{projS} : \Pull\,\vA \to
\MTy$, $\mit{projSeed} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas$, and
$\mit{projStep} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas \to
\Gen\,(\Step\,(\mit{projS}\,\vas)\,\vA)$.  For now, we assume that the following
instance exists:
\[ \instance (\IsSOP\,\vA,\,\{\va : \vA\} \to \IsSOP\,(\vB\,\va)) \RA \IsSOP\,(\Sigma\,\vA\,\vB) \]
Above, $\{\va : \vA\} \to \IsSOP\,(\vB\,\va)$ is a universally quantified
instance constraint; this can be also represented in Agda. The definition of
$\concatMap$ is as follows.
\begin{alignat*}{3}
  & \mathrlap{\concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB }\\
  & \mathrlap{\concatMap\,\{\vA\}\,\{\vB\}\,\vf\,(\Pull\,\vS\,\seed\,\step) = }\\
  & \mathrlap{\ind \Pull\,(\vS,\,\Maybe\,(\Sigma\,\vA\,(\mit{projS} \circ \vf)))\,(({,}) \fmap \seed\,\ap\,\return\,\Nothing) \dlr \lam \case }\\
  & \mathrlap{\ind \ind (\vs,\,\Nothing) \to \step\,\vs \bind \lam \case }\\
  & \ind \ind \ind \ind \Stop                 &&\to \return \Stop\\
  & \ind \ind \ind \ind \Skip\,\vs            &&\to \return \dlr \Skip\,(\vs,\, \Nothing)\\
  & \ind \ind \ind \ind \Yield\,\va\,\vs      &&\to \mdo \{\vs' \fro \mit{projSeed}\,(\vf\,\va); \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\}\\
  & \mathrlap{\ind \ind (\vs,\,\Just\,(\va,\,\vs')) \to \mit{projStep}\,(\vf\,\va)\,\vs' \bind \lam \case }\\
  & \ind \ind \ind \ind \Stop                &&\to \return \dlr \Skip\,(\vs,\,\Nothing) \\
  & \ind \ind \ind \ind \Skip\,\vs'          &&\to \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\\
  & \ind \ind \ind \ind \Yield\,\vb\,\vs'    &&\to \return \dlr \Yield\,\vb\,(\vs,\, \Just\,(\va,\,\vs'))
\end{alignat*}
Here, $\Nothing$ marks the states where we are in the ``outer'' loop, running
the $\Pull\,\vA$ stream until we get its next value. $\Just$ marks the states of
the ``inner'' loop, where we have a concrete $\va : \vA$ value and we run the
$(\vf\,\va)$ stream until it stops. In the inner loop, the machine state type
depends on the $\va : \vA$ value, hence the need for $\Sigma$. How do we get
$\IsSOP$ for $\Sigma$? The key observations are:
\begin{itemize}
 \item Metaprograms cannot inspect the structure of object terms.
 \item Object types do not depend on object terms.
\end{itemize}
Hence, we expect that during unstaging, every $\vf : \Lift\,\vA \to \VTy$
function has to be constant. More generally in the semantics in Appendix \ref{appendix:A}, every function whose domain is a product
of object types and whose codomain is a constant presheaf, is a constant
function. We may call these constancy statements \emph{generativity axioms},
since they reflect the inability of metaprograms to inspect terms, and
``generativity'' often refers to this property in the staged compilation
literature.

Let us write $\Uprod = \List\,\VTy$ and $\Elprod : \Uprod \to \MTy$ for a
universe of finite products of value types. Concretely, in CFTT and the Agda
implementation, we assume the following:
\\\\
\noindent \textbf{Axiom (generativity).}\textit{ For every $\vf : \Elprod\,\vA \to \USOP$ and $\va,\,\va' : \Elprod\,\vA$,
we have that $\vf\,\va = \vf\,\va'$.}
\\\\
See Section \ref{appendix:generativity} in Appendix $\ref{appendix:A}$ for a
validation of this axiom. We remark that there is no risk of staging getting
stuck, because propositional equality proofs get erased, as we described in
Section \ref{sec:semantics-of-staging}. We also remark that although
generativity is inconsistent with inspecting the structure of object terms, it
is consistent with inspecting the structure of object types.

@@

Nonetheless, it seems to be possible and useful to have a
system which simultaneously supports intensional analysis through contextual
types and the 2LTT-style implicit contexts, and it would be interesting
to consider in future work.

@@

We believe that a programming language in the style of this paper would be
highly useful, especially in high-performance functional programming and
domain-specific programming. In particular, the pipeline for using and compiling
monads could look like the following.
\begin{itemize}
\item Users write definitions in a style similar to Haskell, without quotes,
      splices, $\mup$-s and $\mdown$-s, only marking stages in type annotations and
      in let-definitions, with $:=$ and $=$.  Storing monadic actions at runtime
      requires an explicit boxing operation that yields a closure type.
\item Type- and stage-directed elaboration adds the missing operations, and also desugars
      case splitting using $\join$ and the underlying splitting function.
\item In the downstream compiler, fast \& conservative optimization passes are
      sufficient, since abstraction overheads are already eliminated by staging. Eliminating
      dead code and unused arguments would be important, as we have not yet addressed
      this through staging.
\item A systematic way to de-duplicate code would be also needed, probably already
      during staging.
\end{itemize}

@@

\subsection{Generativity} \label{appendix:generativity}
